
\section{Algorithm}\label{sec:background}
%TODO: Johannes

 In this section, we define the topological sort problem and contrast it to Breadth-First-Search (BFS) and Depth-First-Search (DFS).
 Furthermore, we introduce the basics of the parallel algorithm we use.
 \\
 

\mypar{Topological sorting}
 A directed acyclic graph (DAG) describes a partial order. A topological sorting is a total order on a DAG. Let $G = (V,E)$ be a DAG where $V$ is the set of vertices and $E$ is the set of edges. Given $G$, a topological sorting is formally a function $ord: V \longrightarrow \{1,...,n\}$ where $n = |V|$ such that $\forall (v,w) \in E: ord(v) < ord(w)$ \cite[Chapter~9.1]{ottmann2012algorithmen}.

Figure \ref{fig:ts-example} shows a DAG for which one possible topological sorting is A,I,F,B,E,D,H,G,C. Every DAG has at least one topological sorting, but in general there are many different topological sortings for the same graph.

\begin{figure}[!hbp]
 \centering
  \begin{tikzpicture}[->,>=stealth',auto,node distance=1cm,
                    thick,main
                    node/.style={circle,draw,font=\sffamily\scriptsize},text node/.style={draw=none,font=\sffamily\tiny}]

  \node[main node] (1) [draw=black!80,text=black!80] {A};
  \node[main node] (3) [right of=1, node distance=1.2cm]{B};
  \node[main node] (7) [below of=3, node distance=1.2cm] {E};
  \node[main node] (4) [left of=7, node distance=1.2cm] {D};
  \node[main node] (5) [below of=7, node distance=1.2cm] {H};
  \node[main node] (6) [left of=4, node distance=1.2cm] {C};
  \node[main node] (8) [left of=5, node distance=1.2cm] {G};
  \node[main node] (9) [below of=8, node distance=1.2cm] {I};
  \node[main node] (2) [left of=8, node distance=1.2cm] {F};


  \path[every node/.style={font=\sffamily\small}]
    (1) edge (3)
    (3) edge (7)
    (7) edge (4)
    (4) edge (6)
    (7) edge (5)
    (5) edge (8)
    (8) edge (6)
    (9) edge (5)
    (9) edge (2)
    (2) edge (8)
    ;
\end{tikzpicture}

\caption{Example DAG where A,I,F,B,E,D,H,G,C is one possible topological sorting}
\label{fig:ts-example}
\end{figure}

Although the results of the breadth-first-search (BFS) algorithm seem to be similar to topological sortings, they are not equivalent. A topological sorting is a total order with respect to the partial order represented by the input graph. In contrast, the traversal sequences of a BFS (and DFS) generally don’t correspond to a total order induced by a DAG. 
In particular, each node only has to be visited once in BFS - an assumption that does not hold for topological sorting. As a consequence, many of the ideas that are used in parallel BFS (such as the ones presented in \cite{bulucc2011parallel} and \cite{beamer2013direction}) cannot be directly transfered to topological sorting. \\
A simple example is shown in figure \ref{fig:diff-bfs}. The visiting sequence A,C,B is a possible traversal sequence in BFS but is not a topological sorting. The only valid topological sorting for this graph is A,B,C.


\begin{figure}[!hbp]
\centering
 
  \begin{tikzpicture}[->,>=stealth',auto,node distance=1cm,
                    thick,main
                    node/.style={circle,draw,font=\sffamily\scriptsize},text node/.style={draw=none,font=\sffamily\tiny}]

  \node[main node] (1) [draw=black!80,text=black!80] {A};
  \node[main node] (2) [below right of=1, node distance=1.2cm]{B};
  \node[main node] (3) [below of=1, node distance=1.2cm] {C};
 

  \path[every node/.style={font=\sffamily\small}]
    (1) edge (2)
    (2) edge (3)
    (1) edge (3)
    ;
\end{tikzpicture}

\caption{A,C,B is a valid sequence in breadth-first-search (BFS) and depth-first-search (DFS) but not a topological sorting}
\label{fig:diff-bfs}
\end{figure}

Topological sorting has an asymptotic complexity of $\mathcal{O}(|V|+|E|)$ \cite[Chapter~22.4]{cormen2001introduction}. Sequential algorithms have been published by Kahn \cite{kahn1962topological} and by Tarjan  \cite{tarjan1976edge}. The latter is based on a DFS with backtracking.

 \mypar{Parallel algorithm}
 In 1983 M. C. Er \cite{er1983parallel} came up with a parallel approach to retrieve a topological sorting. The algorithm works in 5 steps:
 \begin{enumerate}
        \item Build the graph from the given partial order (optional if the problem is already stated as a graph).
        \item Add a special node value to every node and initialize it to zero
        \item Visit all source nodes (nodes with an indegree of zero) and set their node values to one.
        \item Let all source nodes be front nodes. For all child nodes of the front nodes, proceed in parallel as follows: Let $N_f$ be the node value of a front node and $N_c$ the node value of a child node. If $N_f \leq N_c$, the value of the child is set to $N_f + 1$. After all child nodes have been processed, denote all child nodes as the new front nodes. Repeat this step until there are no further child nodes. 
        \item List all the nodes in ascending order of node values.
 \end{enumerate}
 
To avoid a race condition by concurrent writes to a node’s value, M. C. Er proposes a synchronization after every iteration of step 4. Therefore, any two threads would write the same number to a node’s value, because the barrier ensures that all threads are in the same iteration. This comes at the price of a lower performance, but avoids locking the node's value. However, node values might be rewritten by multiple threads that "follow" each other. For example, in figure \ref{fig:ts-example}, nodes H, G and C could receive the values 2,3,4 initially, but would eventually be overwritten with the values 4,5,6

The asymptotic parallel runtime of the above algorithm is stated by M. C. Er as $\mathcal{O}(D_{max})$, where $D_{max}$ is defined as the maximum distance between a source node and a sink node. This runtime is hard to achieve in practice if one implements step 5 of the algorithm via sorting the nodes with respect to their values. It is not mentioned by M. C. Er how to create the result list of step 5.

\mypar{Improvements}
We propose not to use node values, but to directly put the node into the solution list. This avoids sorting the nodes by their value at the end. However, it must be made sure that no node is written more than once to the result list and race conditions while writing to the solution list must be avoided.

Furthermore, we introduce a \emph{parent counter} to address the problem of several threads processing the same nodes and ``following'' each other throught the graph. For each node, the parent counter is initialized to the number of parent nodes. During the algorithm each thread arriving at a node will decrease the counter by one. It will only process the node if the parent counter is zero. Thus a node will be processed only by the last arriving thread. Care has to be taken of the race condition while updating the parent counter.



 
 
 
 \begin{invisible}
 % Serial TS
 \mypar{Topological sorting}
 \begin{itemize}
  \item What is topological sort, difference to BFS
  \item Input: A set of dependencies (aka partial orders) of the form A $\rightarrow$ B ``A must come before B''
  \item Output: A sequence (aka total order) containing all nodes exactly once. All partial orders must be kept.
  \item Solution not unique
  \item Minimal Example: A->B, A->C, B->C. Valid BFS traversal order: A, C, B. Invalid for TS.
  \item TS can (serially) be solved with Kahn's algorithm \cite{kahn1962topological} or DFS and Backpropagation (Tarjan \cite{tarjan1976edge}). % See Wikipedia
        Note that TS is not equivalent to DFS, e.g. for A->B, B->C, A->D, D->E, DFS and Backpropagation yields A, B, C, D, E, but another valid TS is A, B, D, C, E
  \item Asymptotic runtime: O(|V| + |E|)
As an aside, don't talk about "the complexity of the algorithm.'' It's incorrect,
problems have a complexity, not algorithms.  
 \end{itemize}

 % Multithreaded TS
 \mypar{Parallel algorithm}
 \begin{itemize}
  \item Short overview over algorithm of MC Er
  \item Parallelization over child nodes
  \item His idea with barrier in each step such that even if the index is written by multiple threads, they write the same number => Avoid race condition at writing the index
  \item Our idea: Instead of writing an index, directly write to solution list. As a consequence, we have to make sure that node is written to solution only once. And of course there is a race condition on writing to solution list.
  \item Our idea: First, count (in parallel) how many parents each node has. Each time a node is visited, decrement counter and only write to solution if counter is zero. Of course, there is a race condition on the parent counter.
  \item 3 synchronization points (that is, bottlenecks): 1. Barrier after each level, 2. Lock solution list for appending new nodes, 3. Lock parent counter for decrementing it and checking if it is zero.
  \item Cost
 \end{itemize}

 
\end{invisible}
